---
title: "Clinical Data Analysis"
author: "ChangShen"
date: "10/31/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE,error = TRUE)
knitr::opts_knit$set(root.dir="/Users/dianshen/Desktop/SAS/R Project/Data for R Project")
options(repos =  "http://cran.us.r-project.org")
if(!"pacman" %in%installed.packages()) install.packages("pacman")
library(pacman)
p_load(dplyr,tidyr,plyr,readr,lubridate,Hmisc,knitr,ggplot2, gridExtra)
```



```{r read  files}
## read in the csv files that you need
#>Note: change the path to your local path that contains your .csv files, make sure you data is sorted by name ascending
myfiles = list.files(path=getwd(), pattern="*.csv", full.names=TRUE)
myfilenames<-c("CohortCW","IDiagnosisCW","InpatientVisits","LabCW","MainPatientFile","MedicationCW","Medications","ODiagnosisCW","OutpatientLabs","OutpatientVisits")
for(i in 1:length(myfiles))
{
  filename=myfiles[i];
  data=read_csv(file = filename)
  assign(x = myfilenames[i],value = data)
}
rm(data)
```



```{r}
## functions: transfer a SAS date data in to R .Date format
DateTransfer <- function(x){format(as.Date(x, origin="1960-01-01"))}
```


	
```{r connect files}
## connect the files
#Connect the linkage files with the main files to produce more meaningful results.

##---------------------------------##
## Mainpatient File data cleaning  ##
##---------------------------------##
#1. Create Birthdate&RecordDate by combining month/day/year and transfer to Date format
#2. Caluculate age at record from birthdate and recorddate(round to 2 decimal)
#3. Create catgorical variable Agecat(older(1)>=50/younger(0)<50)
#4. Connect CohortCW to determine unique ID for each observations.
#5. Only keep the most recent observarion for each Unique ID.
#6. Keep the neccessary variables and arrange by Unique ID
#---------------------------------------------------------------------------------------------

MainPatientFile <- MainPatientFile%>%unite("BirthDate",c("month_birth","day_birth","year_birth"),sep="/")%>%
  mutate(year=record_year)%>%
  unite("RecordDate",c("record_month","record_day","record_year"),sep="/")%>%
  mutate(BirthDate=as.Date(BirthDate,format="%m/%d/%Y"),
         RecordDate=as.Date(RecordDate,format="%m/%d/%Y"),
         Age = round(as.numeric(difftime(RecordDate,BirthDate))/365.25,2),
         Agecat = ifelse(Age>=50,1,0)
         )%>%
  mutate(BirthDate = format(BirthDate,"%m/%d/%Y"))%>%
  left_join(CohortCW,by=c("siteID","SiteNo"))%>%
  group_by(uniqueID)%>%arrange(RecordDate)%>%filter(row_number()==n())%>%
  select(uniqueID, sex,BirthDate, race, marital_status, income_bracket, Age, Agecat)%>%arrange(uniqueID)

##----------------------------------------##
## OutpatientLabs File data manipulation  ##
##----------------------------------------##
#1. Date Transformation to R Date format
#2. Left Join Lab_code and CohortCW dataset to OutpatientLabs
#3. Transfer numeric lab result to numeric format(note missing numeric lab result will be transfer as 0, 
#will transfer back to NA in next chunk )
#----------------------------------------------------------------------------------------------------------
OutpatientLabs <- OutpatientLabs%>%left_join(LabCW, by="lab_code")%>%left_join(CohortCW[,1:2], by = "siteID")%>%mutate(
  labdate = DateTransfer(labdate),
  numeric_lab_result = as.numeric(numeric_lab_result)
)

##-----------------------------------------##
## OutpatientVisit File data manipulation  ##
##-----------------------------------------##
#1. Date Transformation to R Date format
#2. Left Join ODiagnosisCW and CohortCW dataset to OutpatientLabs
#----------------------------------------------------------------------------------------------------------
OutpatientVisits <- OutpatientVisits%>%left_join(ODiagnosisCW, by=c("diagnosis_code"="Diagnosis_code"))%>%
  left_join(CohortCW[,1:2], by = "siteID")%>%
  mutate(
  visitdate = DateTransfer(visitdate)
)

##-----------------------------------------##
##  InpatientVisit File data manipulation  ##
##-----------------------------------------##
#1. Date Transformation(admitdate&dischargedate) to R Date format
#2. Left Join IDiagnosisCW and CohortCW dataset to OutpatientLabs
#----------------------------------------------------------------------------------------------------------
InpatientVisits <- InpatientVisits%>%left_join(IDiagnosisCW, by=c("diagnosis_code"="Diagnosis_code"))%>%
  left_join(CohortCW[,1:2], by = "siteID")%>%
  mutate(
  admitdate = DateTransfer(admitdate),
  dischargedate = DateTransfer(dischargedate)
)

##-----------------------------------------##
##    Medications File data manipulation   ##
##-----------------------------------------##
#1. Datetransfer
#2. left join MedicationCW and CohortCW
Medications <- Medications%>%left_join(MedicationCW,by="medication_code")%>%left_join(CohortCW[,1:2], by = "siteID")%>%
  mutate(PrescriptionDate = DateTransfer(PrescriptionDate))

```


```{r A1}
# define A1 for all A1 disease diagnose
# define Unique for all Unique ID
A1 <- c("A1",paste("A1.",1:6,sep=""))
Unique <- data.frame(uniqueID = unique(CohortCW$uniqueID))

#---------------------------------------------------------------------
#Summarize the inpatient visit result of A1 for each unique ID
#Output Data set:
#InpatientVisits_A1
#Variables:
# uniqueID - patient ID unique for each patient
# n_A1-  how many times are they diagnosed A1 prior to 2015
#----------------------------------------------------------------------
InpatientVisits_A1 <- InpatientVisits%>%filter(year(admitdate)<2015)%>%group_by(uniqueID)%>%
  dplyr::summarise(n_A1 = sum(Diagnosis %in% A1),n_Site = n())

#--------------------------------------------------------------------
#Summarize the outpatient visit result of A1 for each unique ID
#Output Data set:
#OutpatientVisits_A1
#Variables:
# uniqueID - patient ID unique for each patient
# n_A1_out- how many times are they diagnosed A1 in OutpatientVisits prior to 2015
#---------------------------------------------------------------------
OutpatientVisits_A1 <- OutpatientVisits%>%filter(year(visitdate)<2015)%>%group_by(uniqueID, visitdate)%>%
  dplyr::summarise(n_A1_out = sum(Diagnosis %in% A1))%>%ungroup()%>%group_by(uniqueID)%>%
  dplyr::summarise(n_A1_out = sum(n_A1_out))

#combined the two data sets and condition on 
DiseaseA <- full_join(InpatientVisits_A1,OutpatientVisits_A1, by='uniqueID')%>%
          right_join(Unique,by='uniqueID')%>%
          mutate(n_A1 = replace_na(n_A1, 0),
                 n_A1_out = replace_na(n_A1_out, 0),
                 A1 = ifelse(n_A1_out>=2 & n_A1>=1,1,0))%>%select(uniqueID, A1)

```

```{r group disease X}
##summarize and find the date for last lab result of each Unique ID and lab name 
OutpatientLabs <- OutpatientLabs%>%group_by(uniqueID,LabName)%>%
  dplyr::summarise(max_labdate=max(labdate))%>%right_join(OutpatientLabs)
#--------------------------------------------------------------------
#Summarize the outpatient visit result of Disease X for each unique ID
#Output Data set:
#ODiseaseX
#Variables:
# uniqueID - patient ID unique for each patient
# PPP- whether the last lab record for ppp is positive
# KOPD- whether the last lab record KOPD>2
#---------------------------------------------------------------------
DiseaseX <- OutpatientLabs%>%
  mutate(PPP = ifelse((LabName=="PPP")&(max_labdate==labdate)&(qual_results=="Positive"),1,0),
         KOPD = ifelse((LabName=="KOPD")&(numeric_lab_result>2),1,0))%>%group_by(uniqueID)%>%
  dplyr::summarise(DiseaseX = as.logical(sum(PPP))|as.logical(sum(KOPD)))%>%
  right_join(Unique,by='uniqueID')%>%mutate(DiseaseX = replace_na(DiseaseX, 0))
```

```{r WE}
Medication <- Medications%>%
  filter(year(PrescriptionDate) %in% c(2015,2016) & `Medication Name`=="WE")


#--------------------------------------------------------------------------------------
#Summarize the outpatient visit result of Disease X for each unique ID
#Output Data set:
# Medications_WE
#Variables:
# uniqueID - patient ID unique for each patient
# duration- prescription of medication duration for each paitient during 2015 and 2016
# WE_6_M- indicator for those who have prescription duration more than 180 days during 
# 2015 and 2016 
#--------------------------------------------------------------------------------------
Medications_WE <-Medication%>%group_by(uniqueID)%>%
  dplyr::summarise(duration = sum(dayssupply*(Refills+1)))%>%right_join(Unique)%>%
  mutate(duration = replace_na(duration, 0),
        WE_6_M =as.numeric(duration>180))%>%select(uniqueID, WE_6_M)

# add WE indicator to show whether the paitient had a WE prescription during 2015 and 2016 and 
#diagnosed A1 prior to 2015
DiseaseA <- DiseaseA%>%mutate(
  WE_indicator = ifelse((uniqueID %in% Medication$uniqueID)&(A1==1),1,0)
)
```


```{r formatdata}
#--------------------------------------------------------------------------------------
#merge 4 datasets we created before to produce the FINALDATA
#Output Data set:
#FINALDATA
#Variables:
# uniqueID - patient ID unique for each patient
# Sex - 
# Birthdate- with format(mm/dd/yyyy)
# race - 
# martital status
# income_braket
# Age 
# Agecat - age category(younger/older)
# A1 - Indicator of whether an individual has chronic condition A1 prior to 2015
# DiseaseX- Indicator of whether an individual has been diagnosed with disease X
# WE_indicator - Indicator of receipt of prescription WE any time during 2015 and 2016 for those with chronic condition A1 prior to 2015 (A1 created above)
# WE_6_M - Indicator of receipt of WE longer than 6 months.
#--------------------------------------------------------------------------------------
FINALDATA <- Reduce(function(...) merge(..., by='uniqueID', all.x=TRUE), list(MainPatientFile,DiseaseA,Medications_WE,DiseaseX))


# Standardize the format of the final main patient data set
FINALDATA <- FINALDATA%>%
  mutate(sex = mapvalues(sex, from = c("0","1"), to = c("Female","Male")),
         race = mapvalues(race, from = as.character(0:5), to = 
                            c("White", "Black", "Asian", "Native American", "Pacific Islander", "Unknown")),
         marital_status = mapvalues(marital_status, from = as.character(1:6), to = 
                                      c("Married", "Divorced", "Separated", "Widowed", "Single", "Other")),
         income_bracket = mapvalues(income_bracket, from = as.character(1:7), to = 
                                      c("<20,000", "20,000 to <40,000", "40,000 to <80,000", 
                                        "80,000 to 120,000", "> 120,000", "Refuse to Respond","Missing")),
         Agecat = mapvalues(Agecat, from = c("0", "1"), to = c("younger", "older"))
  )%>%arrange(uniqueID)

#remove dataset from the environment
rm(IDiagnosisCW, InpatientVisits, InpatientVisits_A1, LabCW, MedicationCW, Medications,ODiagnosisCW, OutpatientLabs, OutpatientVisits, OutpatientVisits_A1, CohortCW, A1)


```

## Data Description
After the database connecting, reformating and creating new summary variable by unique ID(patient),
we get a finaldata includes the personal information, past diagnose(both in and out hospital) and prescription recording.  

#### FINALDATA have `r dim(FINALDATA)[1]` obeservations and `r dim(FINALDATA)[2]` variables

- **uniqueID** - patient ID unique for each patient
- **Sex** - gender for each patient
- **Birthdate**- Birthdat of each patient with format(mm/dd/yyyy)
- **race** - race of paitients including Asian/Black/Native American/Pacific Islander/Unknown/White
- **martital_status** - martital_status of patients(Divorced/Married/Other/Separated/Single/Widowed)
- **income_braket** - income of patients (<20,000/> 120,000/20,000 to <40,000/40,000 to <80,000/80,000 to 120,000/Missing/Refuse to Respond)
- **Age** - age of patient(round to 2 decimals)
- **Agecat** - age category(younger/older)
- **A1** - Indicator of whether an individual has chronic condition A1 prior to 2015
- **DiseaseX** - Indicator of whether an individual has been diagnosed with disease X
- **WE_indicator** - Indicator of receipt of prescription WE any time during 2015 and 2016 for those with chronic condition A1 prior to 2015 (A1 created above)
- **WE_6_M** - Indicator of receipt of WE longer than 6 months.


---

---

#### 0
We can see the detail of the final data(head&tail) as follows.
```{r}
#produce data report
kable(head(FINALDATA,5), caption = "First 5 observations of the FINALDATA")
kable(tail(FINALDATA,5), caption = "Last 5 observations of the FINALDATA")
```

---

#### A

```{r, fig.height = 3.5, fig.width = 5.5, fig.align = "center"}
kable(table(FINALDATA$sex), col.names = c("sex","freqency"), caption ="A table with the frequency distribution of sex in FINALDATA")
kable(table(FINALDATA$race), col.names = c("race","freqency"), caption ="A table with the frequency distribution of race in FINALDATA")
kable(table(FINALDATA$marital_status), col.names = c("marital status","freqency"), caption ="A table with the frequency distribution of marital_status in FINALDATA")
kable(table(FINALDATA$income_bracket), col.names = c("income bracket","freqency"), caption ="A table with the frequency distribution of income_bracket in FINALDATA")
kable(table(FINALDATA$Agecat), col.names = c("age category","freqency"), caption ="A table with the frequency distribution of Agecat in FINALDATA")

plot1 <- ggplot(FINALDATA, aes(x=Age)) +
  geom_histogram(bins = 20,position="identity", color = "white")+ggtitle("Distribution of Age in Final data")+theme_bw();plot1
```

---

#### B
 The proportion of individuals in the cohort had chronic condition A1 prior to 2015 is `r round(sum(FINALDATA$A1)/length(FINALDATA$A1)*100,2)` \%.

---

#### C
```{r}
table_C <- table(FINALDATA$A1,FINALDATA$Agecat)
rownames(table_C) = c("No chronic condition A1","chronic condition A1")
kable(table_C, caption = "age distribution of chronic condition A1 in this population" )
```

---

#### D
proportion of individuals have disease X is `r round(sum(FINALDATA$DiseaseX)/length(FINALDATA$DiseaseX)*100,2)` \%.

---

#### E
```{r}
table_E <- table(FINALDATA$DiseaseX,FINALDATA$sex)
rownames(table_E) = c("No DiseaseX","Have DiseaseX")
kable(table_E, caption = "gender distribution of disease X in the population" )
```

---

#### F
The proportion of individuals classified as having chronic condition A1 prior to 2015,received a prescription for WE (medication name) during the 2015 and 2016 calendar years is `r round(sum(FINALDATA$WE_indicator)/(sum(FINALDATA$A1))*100,2)` \%.

---

#### G
```{r}
table_G <- table(FINALDATA$race,FINALDATA$WE_indicator)
kable(table_G, caption = "the breakdown by race of individuals classified as having chronic condition A1 prior to 2015 and received a prescription for WE (medication name) during the 2015 and 2016 calendar years", col.names = c("Yes(WE& DieaseA1)","No(WE& DieaseA1)") )
```

---

#### H
  proportion of individuals classified as having chronic condition A1 prior to 2015 receiving a prescription for WE (medication name) during the 2015 and 2016 calendar years would be covered longer than 6-month over this time frame is `r round(sum(FINALDATA$WE_6_M&FINALDATA$A1)/(sum(FINALDATA$A1))*100,2)` \%.